Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-05 supplement.tex
add f4 section

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{bm}

@@

$F_{ST}$ can be interpreted as a measure of population structure, given by the
proportion of genetic variance in a multi-population sample that can be
attributed to drift between populations in that sample.

@@

@@

>>>>

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}

@@

$F_{ST}$ can be interpreted as a measure of population structure, given by the
proportion of genetic variance in a multi-population sample that can be
attributed to drift between populations in that sample \cite{Bhatia2013-sq}.
$F_{ST}$ depends on population divergence times as well as subsequent gene flow
and population sizes: migration between populations reduces $F_{ST}$ while
small effective population sizes can lead to large values. For each of the four
inferred demographic models presented here, we computed predicted $F_{ST}$ for
pairs of contemporaneous populations over time. For sampled populations,
predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}). We
computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
pairwise diversity within and between populations as
\begin{equation}
    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
    \label{eq:FST}
\end{equation}

In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
modern humans were much larger than between any pair of modern human
populations or their ancestors (stems). $F_{ST}$ between modern human
populations over the last 100 thousand years remained relatively low, at
$\sim0.1$ or less. In the continuous-migration model, the two stem populations
also were predicted to have low $F_{ST}$ within this range, as migration rates
were inferred to be high enough to maintain genetic similarity. In both the
merger-without-stem-migration and the merger-with-stem-migration, one of the
stem populations was inferred to have a short, severe bottleneck, which led to
increased $F_{ST}$ between stem populations during that time. However, this
period of relatively large $F_{ST}$ did not result in large $F_{ST}$ between
more recent pairs of populations due to their roughly similar levels of
contributions to the ancestors of modern humans. \aprcomment{We caution that
the severe bottleneck may be an artifact of the model parameterization?}

@@

To evaluate the degree to which divergence among modern populations is
inherited from divergence between early stem population, we computed f4
statistics of the form
\begin{equation}
    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
    \E[(f_{\text{stem}_1} - f_{\text{stem}_2})(f_{\text{modern}_1}-f_{\text{modern}_2})]
    \label{eq:f4}
\end{equation}
This statistic measures co-linearity of genetic drift between modern
populations $(f_{\text{modern}_1}-f_{\text{modern}_2})$ and stem populations
$(f_{\text{stem}_1}-f_{\text{stem}_2})$ (see
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
and \citet{Lipson2020-cq}).

To contrast these statistics to divergence across modern populations,
\begin{equation}
    f_2(\text{modern}_1, \text{modern}_2) =
    \E[(f_{\text{modern}_1}-f_{\text{modern}_2})^2]),
\end{equation}
we interpret the $f$-statistics as scalar products between vectors in 
allele frequency space. In particular, the vector 
$v=(f_{\text{modern}_1}-f_{\text{modern}_2})$, 
representing differences in allele frequencies across modern populations,
can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
a component parallel to the divergence $(f_{\text{stem}_1}-f_{\text{stem}_2})$
across stems. We can then decompose 
$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
\begin{equation}
    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
\end{equation}

Finally, the proportion of modern-day structure
$f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
component parallel to ancient population structure is
\begin{equation}
    \alpha^2 = \frac{v_\perp^2}{f_2(\text{modern}_1, \text{modern}_2)} =
    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{modern}_1, \text{modern}_2)}.
\end{equation}
$\alpha^2$ can be interpreted as the proportion of the $f_2$
divergence across modern populations that is aligned with the differences in
allele frequencies across ancient populations. However, drift occurring after
the time of sampling of ancient populations is not correlated to drift
occurring before that time. For this reason, we also interpret $\alpha^2$ as the
proportion of the $f_2$ divergence across modern populations that is explained by
the differences in allele frequencies across ancient populations. We report
this predicted statistic for each of the four inferred demographic models in
Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.75\textwidth]{figures/supp-f4-scenarios}
    \caption{
        \textbf{Present day structure attributable to ancient population structure.}
        The statistic 
        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
        captures how much of present-day population structure can be attributed to 
        population structure among the stems. The blue line represents genetic drift 
        along ancestral branches between modern populations, while the red line 
        represents drift between stem populations. $f_4$ measures the overlap 
        between the blue and red line, which corresponds to shared genetic drift 
        between the two pairs. 
        In scenario A, modern populations are descended from distinct stems, and 
        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
        equals the drift between stems, $f_2(\text{stem}_1, \text{stem}_2)$.
        In scenario B, despite deep population structure, there is no shared drift
        and $f_4=0$. That is, present-day population structure is unrelated to
        structure among stems. In more realistic scenario C, where $\text{modern}_1$
        receives a proportion $m$ of its ancestry from the dotted arrow, the overlap
        is only found along the corresponding path, and $f_4$ depends on both
        ancestry proportion $m$ and drift between stems,
        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)
        = mf_2(\text{stem}_1, \text{stem}_2)$
        \aprcomment{Get better resolution image from Simon}
    }
    \label{fig:supp-f4-scenarios}
\end{figure}

@@

\subsection{Re-inference of an IM model from simulated data}
